---
title: "Exploratory Data Analysis"
author: "Jerry Kiely"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    # number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




\newpage

# Descriptive Statistics

\vspace{1.5em}



## Measures of Central Tendency and Variation

\vspace{1.5em}



### Sources and Types of Data

\vspace{1.5em}



#### Sources of Data

1. Primary Data

* the data is collected by the investigator himself / herself for a specific purpose
* direct method of data collection
* eg. data collected for research through questionnaires, interviews

\vspace{1.5em}

2. Secondary Data

* the data is collected by someone else, but being used by the investigator for some other purpose
* an indirect method of data collection
* eg. census data being used to study the impact of education on income

\vspace{1.5em}

#### Types of Data

1. Structured Data

* information is stored with a high degree of organization
* contains qualitative data, quantitative data, or a mixture of both
* eg. data arranged in an excel file, in rows and columns

\vspace{1.5em}

2. Unstructured Data

* information that either does not have a pre-defined data model and / or is not organized in a pre-defined manner
* eq. emails, tweets, blogs, etc.

\vspace{1.5em}

### Measurement Scales

1. Nominal Scale

* the placing of data into categories without any order or structure
* no numerical relationship between categories - even if numbers are used for representation
* eg. gender, nationality, language, region, etc.

\vspace{1.5em}

2. Ordinal Scale

* the placing of data into categories such that the order of values is meaningful, but relative degree of difference is not known
* eg. ranking the features of a product on a scale of 1 to 5
* the Likert scale - psychometric scale commonly used in questionnaires


| Highly Satisfied | Dissatisfied | Neutral      | Satisfied    | Highly Satisfied | 
|------------------|--------------|--------------|--------------|------------------|
|        1         |       2      |      3       |       4      |         5        |

\vspace{1.5em}

3. Interval Scale

* numeric scale in which the order as well as the relative difference between values is known
* no "true zero"
* eg. temperature can be below $0^\circ C$ 

\vspace{1.5em}

4. Ratio Scale

* numeric scale with an absolute "zero"
* addition, subtraction, multiplication, and division are all valid operations
* eg. height, weight age, etc. - always measured from $0$ to a maximum value




### Measures of Central Tendency

a.k.a. Measures of Central Location

* a single value that describes a set of data by identifying the central position within that set of data

The most commonly used measures of central tendency are: 


* Mean
  * arithmetic mean, commonly known as average
  * sum of all values divided by the number of values 


* Median
  * arrange N data elements in order
  * if N is odd take the middle value
  * if N is even take the average of the two middle values 


* Mode
  * the most frequently occuring value in a data set 


The mean, median, and mode are all valid measures of central tendency, but under different conditions, some 
measures are more appropriate than others. 



\newpage

```{r cars}
summary(cars)
```

## Beyond Mean and Variance




You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
