---
title: "Exploratory Data Analysis"
author: "Jerry Kiely"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```







\newpage





# Data Management

\vspace{1.5em}





## Chapter 1

\vspace{1.5em}

To be completed...

\vspace{1.5em}





\newpage





# Descriptive Statistics

\vspace{1.5em}





## Measures of Central Tendency and Variation

\vspace{1.5em}





### Sources and Types of Data

\vspace{1.5em}



#### Sources of Data

\vspace{1.5em}

1. Primary Data

\vspace{1.5em}

* the data is collected by the investigator himself / herself for a specific purpose
* direct method of data collection
* eg. data collected for research through questionnaires, interviews

\vspace{1.5em}

2. Secondary Data

\vspace{1.5em}

* the data is collected by someone else, but being used by the investigator for some other purpose
* an indirect method of data collection
* eg. census data being used to study the impact of education on income

\vspace{1.5em}



#### Types of Data

\vspace{1.5em}

1. Structured Data

* information is stored with a high degree of organization
* contains qualitative data, quantitative data, or a mixture of both
* eg. data arranged in an excel file, in rows and columns

\vspace{1.5em}

2. Unstructured Data

\vspace{1.5em}

* information that either does not have a pre-defined data model and / or is not organized in a pre-defined manner
* eq. emails, tweets, blogs, etc.

\vspace{1.5em}





### Measurement Scales

\vspace{1.5em}

1. Nominal Scale

\vspace{1.5em}

* the placing of data into categories without any order or structure
* no numerical relationship between categories - even if numbers are used for representation
* eg. gender, nationality, language, region, etc.

\vspace{1.5em}

2. Ordinal Scale

\vspace{1.5em}

* the placing of data into categories such that the order of values is meaningful, but relative degree of difference is not known
* eg. ranking the features of a product on a scale of 1 to 5
* the Likert scale - psychometric scale commonly used in questionnaires

\vspace{1.5em}

| Highly Satisfied | Dissatisfied | Neutral      | Satisfied    | Highly Satisfied | 
|------------------|--------------|--------------|--------------|------------------|
|        1         |       2      |      3       |       4      |         5        |

\vspace{1.5em}

3. Interval Scale

\vspace{1.5em}

* numeric scale in which the order as well as the relative difference between values is known
* no "true zero"
* eg. temperature can be below $0^\circ C$ 

\vspace{1.5em}

4. Ratio Scale

\vspace{1.5em}

* numeric scale with an absolute "zero"
* addition, subtraction, multiplication, and division are all valid operations
* eg. height, weight age, etc. - always measured from $0$ to a maximum value

\vspace{1.5em}





### Measures of Central Tendency

\vspace{1.5em}

a.k.a. Measures of Central Location

\vspace{1.5em}

* a single value that describes a set of data by identifying the central position within that set of data

\vspace{1.5em}

The most commonly used measures of central tendency are: 

\vspace{1.5em}

* Mean
  * arithmetic mean, commonly known as average
  * sum of all values divided by the number of values 

\vspace{1.5em}

* Median
  * arrange N data elements in order
  * if N is odd take the middle value
  * if N is even take the average of the two middle values 

\vspace{1.5em}

* Mode
  * the most frequently occuring value in a data set 

\vspace{1.5em}

The mean, median, and mode are all valid measures of central tendency, but under different conditions, some 
measures are more appropriate than others. 

\vspace{1.5em}

It is recommended to report trimmed mean along with mean when outliers are present. Trimmed mean excludes
extreme data points from the calculation. Typically $5%$ of data from each end is excluded - which will 
give a robust estimate if the underlying distribution is symmetric.

\vspace{1.5em}

| Type of variable             | Best Measure |
|------------------------------|--------------|
| Nominal                      | Mode         |
| Ordinal                      | Median       |
| Interval / Ratio (symmetric) | Mean         |
| Interval / Ratio (skewed)    | Median       |

\vspace{1.5em}

* for a symmetric distribution the mean is appropriate - the mean is at the center
* for a skewed distribution the median is appropriate - the mean is generally not at the center

\vspace{1.5em}





### Measures of Variation

\vspace{1.5em}

a.k.a. Measures of Dispersion

\vspace{1.5em}

* an indication of the spread of measurements
* two datasets can have similar mean but vastly different variability

\vspace{1.5em}

The most commonly used measures of variation are:

\vspace{1.5em}

* Range
  * simple measure of variation
  * difference between highest and lowest values
  * crude measure as it does not take into account all values
  * same units as the original values
* Inter Quartile Range
  * the range between the upper quartile and the lower quartile
  * the quartiles are the values that divide the data into four equal parts
  * the values that divide each part are the first, second, and third quantiles (Q1, Q2, and Q3)
  * same units as the original values
* Variance and Standard Deviation
  * variance is the sum of the squared deviations from the mean divided by the number of data points
  * standard deviation is the positive square root of the variance, and has the same units as the original values

\vspace{1.5em}

### Coefficient of Variation (CV)

\vspace{1.5em}

* relative measure of variation
* used to compare the variability in two data sets
* standard deviation divided by the mean, usually expressed as a percentage
* higher the value of CV, the more variability
* often referred to as the relative standard deviation

\vspace{1.5em}





## Beyond Mean and Variance

\vspace{1.5em}

### Skewness

\vspace{1.5em}

Describes the shape of the data. It is the lack of symmetry.

\vspace{1.5em}

* positively skewed
  * longer right tail
  * mass of distribution concentrated on the left
  * mode < median < mean
* symmetric
  * both tails are equal
  * mass of distribution distributed equally
  * mode = median = mean
* negatively skewed
  * longer left tail
  * mass of distribution concentrated on the right
  * mode > median > mean

\vspace{1.5em}

* if skewness $> 1$ or $< -1$ the distribution is highly skewed
* if skewness is between $1$ and $0.5$ or between $-0.5$ and $-1$ the distribution is moderately skewed
* if skewness is between $0.5$ and $-0.5$ then the distribution is approximately symmetric

\vspace{1.5em}

### Kurtosis

\vspace{1.5em}

Defined as the measure of peakedness. Measured relative to Normal distribution.

\vspace{1.5em}

* mesokurtic
  * Normal distribution
* leptokurtic
  * more acute peak than the Normal distribution
* platykurtic
  * flatter peak than the Normal distribution

\vspace{1.5em}

* kurtosis $= 3$ (excess $= 0$) is called mesokurtic
* kurtosis $< 3$ (excess $< 0$) is called platykurtic
* kurtosis $> 3$ (excess $> 0$) is called leptokurtic

\vspace{1.5em}

### Moments

\vspace{1.5em}

Constants which help us in knowing the characteristics and graphic shape of data.

\vspace{1.5em}





## Bivariate Relationships

\vspace{1.5em}

### Describing a Bivariate Relationships

\vspace{1.5em}

* univariate data - data having one variable
* bivariate data - data having two variables
  * two numeric variables
  * two categorical variables
  * one numeric variable, one categorical variable

\vspace{1.5em}

can be described using:

\vspace{1.5em}

* scatter plot
* correlation coefficient
* simple linear regression

\vspace{1.5em}

### Scatter Plot

\vspace{1.5em}

A scatter plot consists of:

* x-axis - the values of one variable
* y-axis - the values of another variable
* series of dots / observations
* used with two numeric continuous variables

\vspace{1.5em}

interpreting a scatter plot:

\vspace{1.5em}

* positive correlation
* negative correlation
* no correlation


\vspace{1.5em}

```{r data1, echo = FALSE}

data <- read.csv("../../../data/0104_exploratory_data_analysis/live_class/Normality_Assessment_Data.csv", header = TRUE)

```

```{r qqnorm, echo = FALSE}

qqnorm(data$csi, pch = 1, frame = FALSE)
qqline(data$csi, col = "cadetblue", lwd = 2)

```


\vspace{1.5em}

### Pearson's Correlation Coefficient

numerically measures the strength of a linear relation between two variables.

\vspace{1.5em}

\[
-1 \leq r \leq 1
\]

|                      |         |
|----------------------|---------|
| Positive Correlation | $r > 0$ |
| Negative Correlation | $r < 0$ |
| No Correlation       | $r = 0$ |


\vspace{1.5em}

* the two variables can be measured in different units
* not affected by change of origin / scale

\vspace{1.5em}

### Line of Best Fit

\vspace{1.5em}

* a straight line that best represents the data on a scatter plot
* may pass through some, none, or all of the points

\vspace{1.5em}

### Simple Linear Regression

\vspace{1.5em}

The equation of line of best fit used to describe the relationship between two variables

\vspace{1.5em}

mathematical form:

\vspace{1.5em}

\[
  \textbf{Y} = a \textbf{X} + b + e
\]

\vspace{1.5em}

where:

\vspace{1.5em}

* $a$ - the slope
* $b$ - the intercept
* $e$ - the error

\vspace{1.5em}

$a$ and $b$ are population parameters which are estimated using samples.

\vspace{1.5em}

* $\textbf{Y}$ is the dependent variable
* $\textbf{X}$ is the independent variable

\vspace{1.5em}







\newpage







# Data Visualisation

\vspace{1.5em}







## Chapter 1

\vspace{1.5em}

To be completed...

\vspace{1.5em}










\newpage
